Datelife Reports

### 2020-04-25 paper_3-main.Rmd
coming back

---
output:
  pdf_document: default
  html_document: default
---
```{r eval = TRUE, echo = FALSE, warning=FALSE, message=FALSE, results='hide'}
# code to generate the paper data and figures:
# source("data.R")
# inactivated now just for testing purposes
# we are using drake to avoid running the examples everytime we change something in the Rmd and knit
# using this for testing instead:
source("./data/frin_plan.R") # stopped using drake for this, it is better to run the analysis directly
loadd() # loads all the targets listed in the workflow plan of the previous `make()`
library(datelife)
data(opentree_chronograms)
rpackage_available <- FALSE
availability <- ifelse(rpackage_available, "is available", "will be available")
```

\newpage
**abstract.-**
The combination of new analytical techniques, availability of more fossil and molecular
data, and better practices in data sharing has resulted in a steady accumulation
of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
of Life for a large quantity and diversity of organisms in the last few decades. However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication. `datelife` is a service implemented via an R package and a web site
(<http://www.datelife.org/>) for efficient reuse, summary and reanalysis of published data on lineage divergence times. The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string. A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices. Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram. Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query. If there is no information available for any queried taxa, data can be simulated. All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis. Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology. How well this trees work for this purpose still needs to be tested. `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.

**Keywords:** Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations
\newpage
\begin{center}
\textsc{Introduction}
\end{center}
Clade ages represent a fundamental piece of information for evolutionary
understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014]. The primary information needed for these time estimates comes from the fossil record. Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree. That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].

Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics. TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] platforms store and make available published phylogenies and chronograms for easy scientific reuse. All of them can be queried using automatised web procedures, which permit personalized, large scale queries that are also very fast.
OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates). Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branchlengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.

@@

To our knowledge, there is no system allowing the comparison of available source information (apw does this a little bit, but it is a verbal comparison more than a graphical or visual comparison). Other types of technical comparisons: which trees are better (which trees represent reality better), but this is of course a very difficult issue.
 -->
A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
In here we present the formal description and implementation of the `datelife` service, constituted by an R package and a web site (<http://www.datelife.org/>). There is still much room for improvement, and flaws and limitations are addressed below. We strived for the current implementation of `datelife` to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
<!-- as well as tools for comparison of source and summary trees.  -->
<!-- R packages for benchmarking of functionalities and demonstrating services were also developed and made available.>

<!--(BOLD and OToL species names, are they homogeneous/standardized?-->

@@

A `datelife` search currently accepts scientific names only. It can be any named clade or binomial specific. <!--common names (implemented in phylotastic)-->
Chronogram search is performed at the species level, so when input names correspond
to named clades, `datelife` pulls all accepted species names within the
clade from OToL's reference taxonomy to perform the search.
Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
`datelife` processes input names with the taxon name resolution service [TNRS; @Boyle2013],
<!--which "corrects misspelled names and authorities, standardizes variant spellings,
and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
which corrects potentially misspelled names and typos, and standardizes spelling
variations and synonyms <!--to a taxonomic reference of users choosing, default
to otol taxonomy -->, increasing the probability of correctly finding the
queried taxa in `datelife`'s chronogram database.

The chronogram search is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository. Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion. Also, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
<!-- The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence. The Dryad repository stores all types of biological data besides chronograms, and their metadata does not allow differentiating between branch lengths units.  -->
<!-- branch length units that might represent other types of time of divergence that are not proportional to absolute (geologic) time.  -->

Information from source chronograms can be summarised with a summary statistic of tree branch lengths, such as median or mean. A  much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006] is also implemented via the ape package [@Paradis2004].<!-- and the hierarchical clustering method of ttol??-->
The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree. However, we noticed that the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package). Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree. The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to distribute node ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.

@@

test improvement in synthetic otol tree after using bold sequences. Is it really better resolved? Probably not.
Do not include this analysis in paper, maybe in examples, but i think it is not worthy.
-->
In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches. We performed a higher-taxon search to obtain all data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` spp. according to the Open Tree of Life taxonomy). There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig. \ref{fig:schronograms1}).
<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig. \ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche]. -->
Data from these source chronograms was used to generate two types of summary chronograms, median and SDM. As explained in the `Description`, data from source chronograms was first summarised into a single distance matrix (using either the median or the SDM method) and then the available node ages were used as calibrations points over a consensus tree topology, to obtain a dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM. In both cases, ages are generally consistent with source ages.
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->
Different source chronograms often show substantial variation in ages for clades (see, for example, the ongoing debate about crown group age of angiosperms [@barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. For some studies, especially ones based on branch lengths (studies of diversification, timing of events, trait evolution, and more), using a different chronogram may return different results [@title2016macrophylogenies]. Stitching together these chronograms can create a larger tree and uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses still requires more research.

Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms. In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}). Maximum tree ages were only approximately recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
Branch lengths were successfully generated using the BOLD database for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9; Fig. \ref{fig:cvbold}). From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD. Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and details can be consulted in <https://github.com/LunaSare/datelife_examples>.

@@

At the moment, `datelife`'s chronogram database is not very large, storing `r length(opentree_chronograms$trees)` chronograms up to the time the manuscript was written. This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is not open for scientific reuse nor automatised data mining [@Kumar2017]. OToL is the only public tree repository from where `datelife` can currently pull chronograms to construct its database.
A previous version of TimeTree's synthetic chronogram [@Hedges2015] was made available in the OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
<!--
Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
Thus, incorporation of more published chronograms deposited in OpenTree or perhaps pulled from Dryad directly, to `datelife`'s database is crucial to improve its services.
Methods to automatically mine chronogram data from the Dryad repository could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
<!-- What can we do to mine journal databases? -->
<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.

Another potential concern comes from summary chronograms. We currently summarize all source chronogram data by default. Users can subset source data if they have reasons to favor some or one source chronogram over others. Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely. To our knowledge, there is no objective way to determine if an expert chronogram is better than other. Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used. Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations. It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec]. Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require prior assumptions (such as ML methods). This remains to be tested.

Furthermore, even chronograms obtained with primary data can be very different, as observed from the comparison of source chronograms in the Fringillidae example.
A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated]. Comparison of available chronogram data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention.
<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced. However, this is just an assumption.  -->
Characteristics of the data used for dating analyses as well as from the output chronogram itself, such as quality of alignment (missing data, GC content), lineage sampling strategy and proportion, phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and confidence intervals could be used to score quality of source chronograms. To facilitate subsetting of source chronograms following different criteria, this information should be included as metadata manually entered by curators (as is done in OToL) in the future.
Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different. Hence, summarizing chronograms might imply summarizing evolutionary hypothesis. This could be good from certain point of view, since it could help to get a single global evolutionary history for a lineage. But it could also be bad, since we could be loosing part of the evolutionary history that is only being reflected in some chronograms and not from the summary chronogram.
Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history [@antonelli2017supersmart]. This will be implemented in future `datelife` versions.

In other areas of biological research, such as ecology and conservation biology, it has been indicated that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance. Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ [@Webb2008]) for several taxa lacking this information. Making up branch lengths in this or other ways is accepted in scientific publications: @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data. Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods. There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees containing only taxa with molecular or other data and trees that combine those taxa with taxa from taxonomy alone.

>>>>

---
output:
  pdf_document: default
  html_document: default
---

Clade ages represent a fundamental piece of information for evolutionary
understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014]. The primary information needed for these time estimates comes from the fossil record. Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree. That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].

Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics. TreeBASE [@morell1996roots; @Piel2002], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life [OToL; @Hinchliff2015] are platforms that store and make available published phylogenies and chronograms for easy scientific reuse. Tree in all of them can be queried using automatised web procedures, which permit personalized, large scale searches that are also very fast.
OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates). Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branch lengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.

@@

To our knowledge, there is no system allowing the comparison of available source information (apw does this a little bit, but it is a verbal comparison more than a graphical or visual comparison). Other types of technical comparisons: which trees are better (which trees represent reality better), but this is of course a very difficult issue.
 -->
A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
In here we present the formal description and implementation of the `datelife` service, constituted by an R package and a web site (<http://www.datelife.org/>). There is still much room for improvement, and flaws and limitations are addressed below. We strived for the current implementation of `datelife` to perform the basic tasks described above, featuring a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms. 
<!-- , for the benefit of the broader community. -->
<!-- as well as tools for comparison of source and summary trees.  NOT YET u_u-->
<!-- R packages for benchmarking of functionalities and demonstrating services were also developed and made available.>

<!--(BOLD and OToL species names, are they standardized with each other? I think yes-->

@@

A `datelife` search currently accepts scientific names only. It can be any named clade or binomial specific. <!--common names (implemented in phylotastic)-->
The search is performed at the species level, so when input names correspond
to named clades, `datelife` pulls all accepted species names within the
clade from OToL's reference taxonomy and uses all of those.
Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
`datelife` processes input names with the taxon name resolution service [TNRS; @Boyle2013],
<!--which "corrects misspelled names and authorities, standardizes variant spellings,
and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
which corrects potentially misspelled names and typos, and standardizes spelling
variations and synonyms <!--to a taxonomic reference of users choosing, default
to otol taxonomy -->, increasing the probability of correctly finding the
queried taxa in `datelife`'s chronogram database.

Then, a name matching procedure is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository. Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion. Also, the tip names in their stored trees correspond to scientific names at or below the species level. Finally, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence. 
<!-- The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence. The Dryad repository stores all types of biological data besides chronograms, and their metadata does not allow differentiating between branch lengths units.  -->
<!-- branch length units that might represent other types of time of divergence that are not proportional to absolute (geologic) time.  -->

Information from source chronograms can be summarised with a summary statistic of tree branch lengths, such as median or mean. A  much slower, but possibly more accurate Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006] is also implemented via the ape package [@Paradis2004].<!-- and the hierarchical clustering method of ttol??-->
The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree. However, the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package). Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree. The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to fix known node ages and then distribute nodes with unknown ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.

@@

test improvement in synthetic otol tree after using bold sequences. Is it really better resolved? Probably not.
Do not include this analysis in paper, maybe in examples, but i think it is not worthy.
-->
In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches. We performed a higher-taxon search to obtain all data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` spp. according to the Open Tree of Life taxonomy). There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig. \ref{fig:schronograms1}).
<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig. \ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche]. -->
Data from these source chronograms was used to generate two types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms was first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages were used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM. In both cases, ages are generally consistent with source ages, but see next section.
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms. In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}). Maximum tree ages were only approximately recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data from the Barcode of Life Database (BOLD) and branch lengths were successfully generated for all source chronograms. However, dating with PATHd8 (using congruified calibrations) was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}). From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD. Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees can be consulted in <https://github.com/LunaSare/datelife_examples>.

@@

At the moment, `datelife`'s chronogram database is not very large, storing `r length(opentree_chronograms$trees)` chronograms up to the time the manuscript was written. This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, which is not open for scientific reuse nor automatised data mining [@Kumar2017]. OToL is the only public tree repository from where `datelife` can currently pull chronograms to construct its database.
A previous version of TimeTree's synthetic chronogram [@Hedges2015] was made available in the OToL repository, hence the amount of lineages represented in datelife's database is at least as substantial as TimeTree's.
This ensures that some information will be available for any given query, but it does not ensure that the full state of knowledge of time of divergence data will be available for any given lineage.
<!--
Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
Thus, incorporation of more published chronograms deposited in OpenTree, or perhaps directly pulled from the Dryad repository, to `datelife`'s database is crucial to improve its services.
Methods to automatically mine chronogram data from Dryad could be designed and implemented. However, the unit of branch lengths would still need to be determined by hand.
<!-- What can we do to mine journal databases? -->
<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
Consequently, we would like to emphasize on the importance of sharing chronogram data for the scientific community, in repositories that require expert input and manual curation, such as OToL's tree repository.

Another potential concern comes from summary chronograms. We currently summarize by default all source chronograms that overlap with at least two taxa. Users can subset source data if they have reasons to favor some source chronograms over others. Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely. To our knowledge, there is no objective way to determine if an expert chronogram is better than other. Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used. Scientists usually also favor chronograms coming from studies with primary calibrations to ones from secondary calibrations. It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec]. Yet, there are different ways to use secondary calibrations and the bias might not be encountered with other dating methods that do not require setting priors (such as ML methods). In sum, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.

Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
This observation is often encountered in the literature 
<!-- Different source chronograms often show substantial variation in clade ages  -->
(see, for example, the ongoing debate about crown group age of angiosperms [@barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results [@title2016macrophylogenies]. Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.

Summarizing high-quality chronograms might also imply summarizing evolutionary hypothesis. This could be enlightening from certain point of view, since it could help getting a single global evolutionary history for a lineage. It could also be really misleading, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in some chronograms and not from the summary chronogram.
Ideally, we should still rely on time of lineage divergence data obtained from a single analysis using fossil data as primary sources of calibrations, and using fossils that have already been curated as calibrations to date other trees, which should reflect a more homogeneous evolutionary history [@antonelli2017supersmart]. This will be implemented in future `datelife` versions.

<!-- still requires more research. -->
<!-- A large discrepancy in time of lineage divergence across expert knowledge is well known for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].  -->
<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced. However, this is just an assumption.  -->

Alternatively, one could try to choose the "best" chronogram. Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms. Some characteristics that are often cited as a measure of improved quality in dating studies addressing long-studied organisms are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals. To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the near future.

<!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->

In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance. Hence, we allow accepted ways of creating branch lengths in the absence of starting branch length information (such as BLADJ [@Webb2008]) for several taxa lacking this information. Making up branch lengths in this or other ways is accepted in scientific publications: @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data. Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods. There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data and trees that are reconstructed using taxonomy.

### 2020-04-25 paper_drafts/Intro.Rmd
coming back

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    toc: false
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: library_red.bib
csl: systematic-biology.csl
link-citations: yes
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---
`datelife`'s database includes all chronograms derived from expert knowledge, and this includes those derived from secondary calibrations.

>>>>

---
output:
  bookdown::pdf_document2:
    keep_tex: yes
    toc: false
textalign: left  # this does not work...
geometry: margin = 1in
bibliography: ../paper_references.bib
csl: ../systematic-biology.csl
link-citations: yes
linestretch: 2
header-includes:
    - \usepackage[left]{lineno}
    - \linenumbers
nocite: |
  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
---
`datelife`'s database includes all chronograms derived from expert knowledge, and this includes those derived from secondary calibrations.
